String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-05 2023-1-19/sections/algorithms.tex
solving nonemptiness of cefa

Let $\aut = (R, Q, \Sigma, \delta, I, F)$ be a CEFA and $\costset_\aut = \{\myvec{v} \mid \exists q, a, q'.\ (q, a, q', \myvec{v}) \in \delta\}$. 
Moreover, let $\aut'$ be an NFA over the alphabet $\costset_\aut$ that is obtained from $\aut$ by ignoring the characters, that is, the transitions of $\NFA'$ are the tuples $(q, \myvec{v}, q')$ such that $(q, a, q', \myvec{v}) \in \delta$ for some $a \in \Sigma$.
%Then $\cefaout(\aut) = \parikh(\NFA')$.
\begin{observation}\label{obs-cefa-nfa}
For each CEFA $\aut$, $\cefaout(\aut) = \left\{\sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v}  \mid \eta \in \parikh(\NFA') \right\}$.
\end{observation}

From Observation~\ref{obs-cefa-nfa}, we know that the following two conditions are equivalent, 
\begin{itemize}
%\item CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$, namely, 
%there are an assignment $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{1}}, \cdots, \myvec{v_{n}}$ such that $(\theta(x_1); \myvec{v_1}) \in \Lang(\cB_1)$, $\cdots$, $(\theta(x_n); \myvec{v_n}) \in \Lang(\cB_n)$ and $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true,
%
\item there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true, 
%CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ 
%
\item 
there are $\eta_1 \in \parikh(\cB'_1), \cdots, \eta_n \in \parikh(\cB'_n)$ such that 
%
%\begin{equation}\label{eqn-cefa-lia}
\[
\varphi \left[\sum \limits_{\myvec{v} \in \costset_{\cB_1}} \eta(\myvec{v}) \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum \limits_{\myvec{v} \in \costset_{\cB_n}} \eta(\myvec{v}) \myvec{v} \big{/}R_{\cB_n} \right]
\]
%\end{equation} 
%
is true. 
\end{itemize}

Therefore, to solve the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$, it is sufficient to compute the existential LIA formulas $\psi_{\cB'_1}(\fZ_{\costset_{\cB_1}}), \cdots, \psi_{\cB'_n}(\fZ_{\costset_{\cB_n}})$, which define the Parikh images of $\cB'_1$, $\cdots$, $\cB'_n$ respectively, and solve the satisfiability of the following existential LIA formula
\begin{equation}\label{eqn-cefa-lia}
\bigwedge \limits_{i \in [n]} \psi_{\cB'_i}(\fZ_{\costset_{\cB_i}}) \wedge 
\varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right],
\end{equation}
where $\anivarz_{\myvec{v}}$'s are integer variables to represent the number of occurrences of $\myvec{v}$.

@@

We also have the following observation: For each $i \in [n]$ and $\eta \in \parikh(\cB'_i)$, we have 
$$\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v} = \sum \limits_{\myvec{v} \in \costset(\cB_i) \setminus \{\myvec{0}\}} \eta(\myvec{v}) \myvec{v}.$$
In other words, the $\myvec{0}$-labeled transitions in $\cB'_i$ do not contribute to the sum $\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v}$.

Therefore, we can reduce the sizes of $\cB'_1, \cdots, \cB'_n$ by Step II presented in the sequel.

@@

For $i \in [n]$, let  $\cC_i$ denote the DFA obtained from $\cB'_i$ by executing Step 2. From the construction, we know that $\parikh(\cC_i) = \parikh(\cB'_i)$ for each $i \in [n]$.
Therefore, we compute LIA formulas from $\cC_i$'s, instead of $\cB'_i$'s, to represent Parikh images.

\paragraph*{Step 3.} For each $i \in [n]$, compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cB_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then solve the satisfiability of the following existential LIA formula 
%
\[
\bigwedge \limits_{i \in [n]} \psi_{\cC_i}(\fZ_{\costset_{\cB_i}}) \wedge 
\varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\]

Finally, although we propose techniques to reduce the sizes of $\cB'_i$'s, it is still relatively expensive to compute LIA formulas from NFA and solve their satisfiability, especially for those problem instances with positive answers, namely, those instances where CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty w.r.t. $\varphi$.   
Therefore, in the sequel, we propose a heuristic to search for solutions for those positive problem instances by utilizing under approximations, before Step 3. Intuitively, we search for solutions by bounding the lengths of the strings that are accepted by $\cC_i$'s. The heuristic uses a constant bound $B > 0$. Since the heuristic is applied before Step 3, we call it Step 2.5, to denote the fact that it is applied after Step 2 and before Step 3.

For $i \in [n]$, suppose $R_{\cB_i} = (r_{i,1},\cdots, r_{i, k_i})$ and $\cC_i = (Q'_i, \costset_{\cB_i}, \delta'_i, q_{i,0}, F'_i)$.

\paragraph*{Step 2.5.} Initially, let $len := 0$, and for each $i \in [n]$, let $\confset_{\cC_i}:=\{(q_{i,0}, \myvec{0})\}$ and $\psi_{\cC_i}:= \lfalse$.  Then iterate the following procedure until $len = B$.
\begin{quote}
For each $i \in [n]$, let $\confset'_{\cC_i}:=\confset_{\cC_i}$ and do the following:  For each $(q, \myvec{v}) \in \confset_{\cC_i}$,
\begin{itemize}
\item if $q \in F'_i$, then let $\psi_{\cC_i}:=\psi_{\cC_i} \vee (\bigwedge \limits_{j \in [k_i]} r_{i,j} = v_j)$, 
%
\item let $\confset'_{\cC_i}:=\confset'_{\cC_i} \setminus \{(q, \myvec{v})\}$, 
%
\item for each transition $(q, \myvec{v'}, q') \in \delta'_i$, let $\confset'_{\cC_i}:=\confset'_{\cC_i} \cup \{(q', \myvec{v}+ \myvec{v'})\}$. 
\end{itemize}
Let $\varphi:=\varphi \wedge \psi_{\cC_i}$.

If $\varphi$ is satisfiable, then return $\ltrue$. Let $len: = len+1$. \qed
\end{quote}

%Furthermore, for each $i \in [n]$, we can remove the redundancy of characters/vectors in $\cC_i$.

@@

}
%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%
%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%

@@

\begin{algorithm}[h]
  \caption{High-level algorithm}
  \label{alg:high}
  \begin{algorithmic}[1]
    \Require Conjunction of literals $\varphi$ and conjunction of linear literals $\psi$
    \Ensure \emph{sat} or \emph{unsat}
    \Statex
    \State $finalAuts \leftarrow \emptyset$
    \ForAll{string variables $x$ occurring in $\varphi$}
    \State $\aut_{len} \leftarrow$ intersection of all pre-images of $i=|x|$ in $\varphi$
    \State $\aut_{regex} \leftarrow$ intersection of all CEFAs of $x \in \regex$ in $\varphi$
    \State $\aut_x \leftarrow$ intersection of $\aut_{len}$ and $\aut_{regex}$
    \State $\aut_{nfa} \leftarrow$ NFA form of $\aut_x$
    \If{$\aut_{nfa}$ is empty}
    \State \textbf{return} \emph{unsat}
    \EndIf
    \State $finalAuts \leftarrow finalAuts \cup \{\aut_x\}$
    \EndFor
    \If{$\algfun{isEmpty}(finalAuts, \psi)$ }
    \State \textbf{return} \emph{unsat}
    \Else
    \State \textbf{return} \emph{sat}
    \EndIf
  \end{algorithmic}
\end{algorithm}

>>>>

Let $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$ be a CEFA and $\costset_\aut = \{\myvec{v} \mid \exists q, a, q'.\ (q, a, q', \myvec{v}) \in \delta\}$. 
Moreover, let $\aut' = (Q, \costset_\aut, \delta', I, F)$ be an NFA over the alphabet $\costset_\aut$ that is obtained from $\aut$ by dropping the accepting condition and ignoring the characters, that is, $\delta'$, the transition relation of $\aut'$, comprises the tuples $(q, \myvec{v}, q')$ such that $(q, a, q', \myvec{v}) \in \delta$ for some $a \in \Sigma$.
%Then $\cefaout(\aut) = \parikh(\NFA')$.
%
\begin{observation}\label{obs-cefa-nfa}
For each CEFA $\aut = (R, Q, \Sigma, \delta, I, F, \alpha)$, 
%
$$\cefaout(\aut) = \left\{\sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \  \big{|}\ \eta \in \parikh(\NFA') \mbox{ and }  \alpha\left[ \sum \limits_{\myvec{v} \in \costset_\aut} \eta(\myvec{v}) \myvec{v} \big{/} R \right] \mbox{ is true}\right\}.$$
%
\end{observation}

For $i \in [n]$, let $\alpha_i$ be the accepting condition of $\cB_i$. 
Then from Observation~\ref{obs-cefa-nfa}, we know that the following two conditions are equivalent, 
\begin{itemize}
%\item CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$, namely, 
%there are an assignment $\theta: \{x_1, \cdots, x_n\} \rightarrow \Sigma^*$ and vectors $\myvec{v_{1}}, \cdots, \myvec{v_{n}}$ such that $(\theta(x_1); \myvec{v_1}) \in \Lang(\cB_1)$, $\cdots$, $(\theta(x_n); \myvec{v_n}) \in \Lang(\cB_n)$ and $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true,
%
\item there are $\myvec{v_{1}} \in \cefaout(\cB_1), \cdots, \myvec{v_{n}} \in \cefaout(\cB_n)$ such that $\varphi[\myvec{v_{1}}/R_{\cB_{1}}, \cdots, \myvec{v_{n}}/R_{\cB_{n}}]$ is true, 
%CEFA in $\Lambda'_{x_1} = \{\cB_1\}, \cdots, \Lambda'_{x_n} = \{\cB_n\}$ are nonempty w.r.t. $\varphi$ 
%
\item 
there are $\eta_1 \in \parikh(\cB'_1), \cdots, \eta_n \in \parikh(\cB'_n)$ such that 
%
%\begin{equation}\label{eqn-cefa-lia}
\[
\bigwedge \limits_{i \in [n]} \alpha_i\left[ \sum \limits_{\myvec{v} \in \costset_{\cB_i}} \eta_i(\myvec{v}) \myvec{v} \big{/}R_{\cB_i}  \right] \wedge \varphi \left[\sum \limits_{\myvec{v} \in \costset_{\cB_1}} \eta_1(\myvec{v}) \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum \limits_{\myvec{v} \in \costset_{\cB_n}} \eta_n(\myvec{v}) \myvec{v} \big{/}R_{\cB_n} \right]
\]
%\end{equation} 
%
is true. 
\end{itemize}

Therefore, to solve the nonemptiness of CEFA in $\Lambda'_{x_1}, \cdots, \Lambda'_{x_n}$ w.r.t. $\varphi$, it is sufficient to compute the existential LIA formulas $\psi_{\cB'_1}(\fZ_{\costset_{\cB_1}}), \cdots, \psi_{\cB'_n}(\fZ_{\costset_{\cB_n}})$ to represent the Parikh images of $\cB'_1$, $\cdots$, $\cB'_n$ respectively, where $\fZ_{\costset_{\cB_i}} = \{\anivarz_{i,\myvec{v}} \mid \myvec{v} \in \costset_{\cB_i}\}$ for $i \in [n]$, and solve the satisfiability of the following existential LIA formula
%
%\begin{equation}\label{eqn-cefa-lia}
\[
\begin{array}{l}
%\scriptsize
\bigwedge \limits_{i \in [n]} \left(\psi_{\cB'_i}(\fZ_{\costset_{\cB_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cB_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right)\  \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
%\end{equation}
Intuitively, the integer variables $\anivarz_{i, \myvec{v}}$ represent the number of occurrences of $\myvec{v}$ in the strings accepted by $\cB'_i$.
%where $\anivarz_{\myvec{v}}$'s are integer variables to represent the number of occurrences of $\myvec{v}$.

@@

Furthermore, we have the following observation: For each $i \in [n]$ and $\eta \in \parikh(\cB'_i)$, we have 
$\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v} = \sum \limits_{\myvec{v} \in \costset(\cB_i) \setminus \{\myvec{0}\}} \eta(\myvec{v}) \myvec{v}.$
In other words, the $\myvec{0}$-labeled transitions in $\cB'_i$ do not contribute to the sum $\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v}$.

Therefore, we can reduce the sizes of $\cB'_1, \cdots, \cB'_n$ by Step II in the sequel.

@@

For $i \in [n]$, let  $\cC_i$ denote the DFA obtained from $\cB'_i$ by executing Step 2 and $\costset_{\cC_i} := \costset_{\cB_i} \setminus \{\myvec{0}\}$. From the construction, we know that $\parikh(\cC_i) = \prj_{\costset_{\cC_i}}(\parikh(\cB'_i))$ for each $i \in [n]$.
Therefore, we compute LIA formulas from $\cC_i$'s, instead of $\cB'_i$'s, to represent the Parikh images.

\paragraph*{Step 3.} For each $i \in [n]$, compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cC_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then solve the satisfiability of the following existential LIA formula 
%
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]

%\[
%\bigwedge \limits_{i \in [n]} \psi_{\cC_i}(\fZ_{\costset_{\cB_i}}) \wedge 
%\varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
%\]

Finally, although we propose techniques to reduce the sizes of $\cB'_i$'s, it is still relatively expensive to compute LIA formulas from $\cC_i$'s and solve their satisfiability, especially for those problem instances with positive answers, namely, those instances where CEFA in $\Lambda_{x_1}, \cdots, \Lambda_{x_n}$ are nonempty w.r.t. $\varphi$.   
Therefore, in the sequel, we propose a heuristic to search for solutions for those positive problem instances by utilizing under approximations, to be executed before Step 3. Intuitively, we search for solutions by bounding the lengths of the strings that are accepted by $\cC_i$'s. The heuristic uses a constant bound $B > 0$ and searches for solutions incrementally, starting from strings of length $0$. Since the heuristic is executed after Step 2 and before Step 3, we call it Step 2.5.

For $i \in [n]$, suppose $R_{\cB_i} = (r_{i,1},\cdots, r_{i, k_i})$ and $\cC_i = (Q'_i, \costset_{\cC_i}, \delta'_i, q_{i,0}, F'_i)$.

\paragraph*{Step 2.5.} Initially, let $len := 0$, and for each $i \in [n]$, let $\confset_{\cC_i}:=\emptyset$, $\confset'_{\cC_i}:=\{(q_{i,0}, \myvec{0})\}$, and $\psi_{\cC_i}:= \lfalse$.  Then iterate the following procedure, until $len = B+1$.
\begin{enumerate}
\item Let $\xi: = \bigwedge \limits_{i \in [n]} \alpha_i \wedge \varphi$.
%
\item For each $i \in [n]$, do the following.
\begin{enumerate}
\item Let $\confset_{\cC_i}:= \confset_{\cC_i} \cup \confset'_{\cC_i}$, $temp:=\confset'_{\cC_i}$, $\confset'_{\cC_i}:=\emptyset$.
%
\item For each $(q, \myvec{v}) \in temp$,
\begin{enumerate}
\item if $q \in F'_i$, then let $\psi_{\cC_i}:=\psi_{\cC_i} \vee (\bigwedge \limits_{j \in [k_i]} r_{i,j} = v_j)$, 
%
%\item let $\confset'_{\cC_i}:=\confset'_{\cC_i} \setminus \{(q, \myvec{v})\}$, 
%
\item for each $(q, \myvec{v'}, q') \in \delta'_i$, let $\confset'_{\cC_i}:=\confset'_{\cC_i} \cup \{(q', \myvec{v}+ \myvec{v'})\}$. 
\end{enumerate}
%
%\item 
%
\item Let $\xi := \xi \wedge \psi_{\cC_i}$. 
\end{enumerate}
%
\item If $\xi$ is satisfiable, then report $\ltrue$ and the whole algorithm terminates. 
%
\item Let $len: = len+1$.
\end{enumerate}
Intuitively, for each $i \in [n]$, $\confset'_{\cC_i}$ stores the set of ``configurations'' of $\cC_i$ that were newly discovered, and $\confset_{\cC_i}$ stores those that have been discovered before.
%Furthermore, for each $i \in [n]$, we can remove the redundancy of characters/vectors in $\cC_i$.

@@

@@

\begin{algorithm}[h]
  \caption{High-level algorithm}
  \label{alg:high}
  \begin{algorithmic}[1]
    \Require Conjunction of literals $\varphi$ and conjunction of linear literals $\psi$
    \Ensure \emph{sat} or \emph{unsat}
    \Statex
    \State $finalAuts \leftarrow \emptyset$
    \ForAll{string variables $x$ occurring in $\varphi$}
    \State $\aut_{len} \leftarrow$ intersection of all pre-images of $i=|x|$ in $\varphi$
    \State $\aut_{regex} \leftarrow$ intersection of all CEFAs of $x \in \regex$ in $\varphi$
    \State $\aut_x \leftarrow$ intersection of $\aut_{len}$ and $\aut_{regex}$
    \State $\aut_{nfa} \leftarrow$ NFA form of $\aut_x$
    \If{$\aut_{nfa}$ is empty}
    \State \textbf{return} \emph{unsat}
    \EndIf
    \State $finalAuts \leftarrow finalAuts \cup \{\aut_x\}$
    \EndFor
    \If{$\algfun{isEmpty}(finalAuts, \psi)$ }
    \State \textbf{return} \emph{unsat}
    \Else
    \State \textbf{return} \emph{sat}
    \EndIf
  \end{algorithmic}
\end{algorithm}
}
%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%
%%%%%%%%%%%% the original texts by Denghang %%%%%%%%%
